---
title: "Course Project"
author: "Nadezda Gushchina"
date: "8 Sep 2015"
output: html_document
---

#Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible 
to collect a large amount of data about personal activity relatively 
inexpensively. 
These type of devices are part of the quantified self movement – a group of 
enthusiasts who take measurements about themselves regularly to improve their 
health, to find patterns in their behavior, or because they are tech geeks. One 
thing that people regularly do is quantify how much of a particular activity they 
do, but they rarely quantify how well they do it. In this project, your goal will 
be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly 
in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting 
Exercise Dataset). 
The goal of the project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set. 

#Loading the data
Download the train and test data and read into "train" and "test" objects
```{r load}
        download.file(
url="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile= "train.csv", method = "curl")
        train<-read.csv("train.csv",na.strings=c("NA","#DIV/0!",""))
           download.file(
url="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile= "test.csv", method = "curl")
       test<-read.csv("test.csv",na.strings=c("NA","#DIV/0!",""))     
```

#Review and clean the data

First we will look at the data and calculate the number of missing values per 
variable in the training set
```{r clean}
       dim(train)
        a<-colSums(is.na(train)) #calculate the number of missing values per variable
        table(a)
```
As we see 60 variables in the training set have no missing values while the rest have most of the values missing. Therefore we will only select the variables which have no missing data in the dataset. We also remove the first column with the ID.
We will select the same variables in the test set (except classe variable)
```{r cleandata}
        trainclean<-train[,a==0] 
        trainclean<-trainclean[,-1]
        dim(trainclean)
        testclean<-test[,colnames(trainclean)[1:58]]
        dim(testclean)
```
#Split the train data into training and testing samples
Use caret package to partition data into training (60%) and testing (40%) sets.
```{r partitioning}
       library("caret")
       inTrain<-createDataPartition(y=trainclean$classe,p=0.6,list=FALSE)
        training<-trainclean[inTrain,]
        testing<-trainclean[-inTrain,]
```

#Predict with decision tree
```{r tree}
        library(rpart)
        library(rpart.plot)
        library(rattle)
        set.seed(33)
        modelFit <- rpart(classe ~ ., data=training, method="class")
        fancyRpartPlot(modelFit)
        prediction<-predict(modelFit, testing, type = "class")
        confusionMatrix(prediction, testing$classe)
```   

Accuracy with decision tree is 88%. Out of sample error is 12%.

#Predict with random forests
We will now use random forests setting ntree to 100.
```{r random_forests}
        library(randomForest)
        set.seed(33)
        modFit <- train(classe~.,data=training,method="rf",ntree=100)
        prediction<-predict(modFit, testing)
        confusionMatrix(prediction, testing$classe)
```  
The accuracy of random forests is much higher than decision tree (99.96%). 
Out of sample error is 0.04%.
#Predict based on test set
Now we predict based on the test set using the random forest which proved to 
provide higher accuracy.
```{r prediction}
        answers<-predict(modFit, testclean)
```  
#Generate the file for submission
```{r submission}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers)
```